---
title: 'Statistical Inference Course Project - Part 2: Basic Inferential Data Analysis'
author: "Ely Xavier"
date: "March 10, 2018"
output:
   html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This report aims to analyze the ToothGrowth data in the R datasets package. We are going to:

1. Load the ToothGrowth Data and perform some basic EDA
2. Provide a basic summary of the data
3. Use confidence intervals and hypothesis test to compare tooth growth by supp and dose (only use the techniques from class, even if there are other approaches worth considering)
4. State your conclusions and the assumptions needed for your conclusions

## Analysis
Load the required libraries and data

```{r libraries}
library(ggplot2)

#Load ToothGrowth data (The effect of Vitamin C on Tooth Growth in Guinea Pigs)
data(ToothGrowth)

#Display summary of the dataset

summary(ToothGrowth)
```

```{r head}
head(ToothGrowth)
```

```{r str}
str(ToothGrowth)
```

The data has 60 observations and 3 variables (from the str() we get the type of variables): 

1. len (numeric) - Tooth length 
2. supp (factor) - Supplement type (VC or OJ) 
3. dose (numeric) - Dose in milligrams

```{r plots}
qplot(supp,len,data=ToothGrowth, facets=~dose, main="Tooth Growth of Guinea Pigs by Supplement Type and Dosage (in mg)",xlab="Supplement Type", ylab="Tooth Length") + geom_boxplot(aes(fill = supp))
```

There is a positive correlation between the dosage amount (i.e., as the dosage increases the tooth growth increases). For VC, the tooth growth has a linear relationship with dosage. For higher dosage (2.0mg) the rate of tooth growth is not as significant as the OJ supplement. In general, the OJ supplement seems to induce more tooth growth than VC except at higher dosage (2.0 mg).

## Hypothesis Testing

Assumptions:

1. The variables must be independent and identically distributed (iid).
2. Variances of tooth growth are different when using different supplement and dosage.
3. Tooth growth follows a normal distribution.


### Hypothesis for Supplement OJ vs. VC

Let our null hypothesis be: "there is no difference in tooth growth when using the supplement OJ and VC"; length OJ = length VC

Let our alternate hypothesis be: "there are more tooth growth when using supplement OJ than VC"; length OJ > length VC

Here we obtain the tooth growth by supplement type by splitting the data.

```{r split data}
OJ = ToothGrowth$len[ToothGrowth$supp == 'OJ']
VC = ToothGrowth$len[ToothGrowth$supp == 'VC']
```

We will perform the following t-test: One-tailed independent t-test with unequal variance.

```{r t-test1}
t.test(OJ, VC, alternative = "greater", paired = FALSE, var.equal = FALSE, conf.level = 0.95)
```

As the p-value (0.03) is lower than 0.05 (the default value for the tolerance of the alpha error), then, we reject the null hypothesis. That can be interpreted as there is approximately a 3% chance of obtaining an extreme value for the difference in mean of tooth growth.

Based on this p-value, we can conclude that it is very likely that supplement OJ has a more significant effect on tooth growth than supplement VC.

### The Hypothesis for the Dosage

The null hypothesis is: there is no difference in tooth growth between dosages Our alternate hypothesis is: there are more tooth growth when the dosage increases

Extract the tooth growth by dosage.

```{r t-test2}
doseHalf = ToothGrowth$len[ToothGrowth$dose == 0.5]
doseOne = ToothGrowth$len[ToothGrowth$dose == 1]
doseTwo = ToothGrowth$len[ToothGrowth$dose == 2]

# One-tail independent t-test with unequal variance
t.test(doseHalf, doseOne, alternative = "less", paired = FALSE, var.equal = FALSE, conf.level = 0.95)
```

As the p-value (6.342e-08) is lower than 0.05 (the default value for the tolerance of the error alpha), we reject the null hypothesis. That can be interpreted as there is almost null chances of obtain an extreme value for the difference in mean of those dosages (doseHalf < doseOne) on tooth growth.

```{r t-test3}

# One-tail independent t-test with unequal variance
t.test(doseOne, doseTwo, alternative = "less", paired = FALSE, var.equal = FALSE, conf.level = 0.95)
```

The p-value is 9.532e-06, close to 0. Then we reject the null hypothesis. That can be interpreted as there are almost null chances of obtaining an extreme value for the difference in mean of those dosages (doseOne < doseTwo) on the tooth growth. The value is extreme (that is why we reject the null hypothesis)

Finally, based on these low p-values, we can conclude that it is very likely that dosage has a positive correlation; the higher the dosage, the higher the tooth growth.

### The Hypothesis for Supplement OJ vs. V: Dosage = 2.0 mg

From the boxplot, we observed that the tooth growth for supplement OJ and VC is similar at dosage 2.0 mg. To test if it is indeed the case, we will test the following hypothesis:

The null hypothesis is: no difference in tooth growth when using the supplement OJ and VC at dosage 2.0 mg

The alternate hypothesis is: difference in tooth growth when using supplement OJ and VC at dosage 2.0 mg

```{r t-test4}
OJ2 = ToothGrowth$len[ToothGrowth$supp == 'OJ' & ToothGrowth$dose == 2]
VC2 = ToothGrowth$len[ToothGrowth$supp == 'VC' & ToothGrowth$dose == 2]

# Two-tail independent t-test with unequal variance
t.test(OJ2, VC2, alternative = "two.sided", paired = FALSE, var.equal = FALSE, conf.level = 0.95)
```
The p-value (0.964) confirms our suspicion, which we can not reject the null hypothesis (p-value is higher than 0.05 (the default value for the tolerance of the error alpha).

Then, there is insufficient evidence to show that there is a difference in tooth growth when using supplement OJ and VC at dosage 2.0 mg.

## Conclusions

Given the following assumptions:

1) The sample is representative of the population
2) The distribution of the sample means follows the Central Limit Theorem

In reviewing our t-test analysis from above, we can conclude that supplement delivery method does not affect tooth growth. However, increased dosages do result in increased tooth length.